---
title: Validating Automatic Mileage & Cost Data
author: "Michael Wimsatt"
date: "February 21, 2015"
output: html_document
layout: post
published: true
tags: [R, mileage, automatic]
comments: true
---

# Validating Automatic data

*Note: Code for this analysis available [on GitHub](https://github.com/wimsy/mileage_analysis/blob/master/validation.Rmd).*

Last March I purchased [Automatic](http://www.automatic.com/), an adapter that plugs into my car's diagnostic port and syncs with an app on my iPhone to report and record data about my driving. Among other things, Automatic provides:

- route, miles and time driven
- gas consumed
- estimated cost of gas consumed
- gas-wasting events like hard braking and speeds over 70 mph

During the first few months, I had trouble getting the adapter to kick in reliably when a trip started, and it still alerts me regularly that it hasn't connected to the smartphone app. Reliability has certainly improved, but I wanted to get an idea of accuracy -- is the cost and gas mileage being reported actually close to what I'm experiencing? To get some independent data, I started recording odometer readings, gallons and price at each fill-up in August.

So... How well do the two datasets match up?

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(lubridate)
library(plyr)
library(dplyr)
library(ggplot2)
library(zoo)
library(knitr)
library(scales)

# Load trip data

root.dir <- '~/Documents/Projects/mileage/'
tripsfile = paste0(root.dir, 'data//raw_trips.csv')
trips <- read.csv(tripsfile, stringsAsFactors = FALSE)

# Note, I know I didn't leave the Easern time zone during this period 
# with this car.
tz = 'America/New_York'
trips$Start.Time <- ymd_hm(trips$Start.Time, tz = tz)
trips$End.Time <- ymd_hm(trips$End.Time, tz = tz)
trips <- trips[order(trips$Start.Time), ]

# Load mileage data

mileagefile = paste0(root.dir, 'data//mileage.csv')
mileage <- read.csv(mileagefile, stringsAsFactors = FALSE)

# Note, I know I didn't leave the Easern time zone during this period 
# with this car.
tz = 'America/New_York'
mileage$Timestamp <- mdy_hms(mileage$Timestamp, tz = tz)
mileage <- mileage[order(mileage$Timestamp), ]
mileage <- na.omit(mileage)
mileage$diff.miles[2:nrow(mileage)] <- diff(mileage$Miles)

tdf <- trips %>%
    select(Start.Time, 
           Distance..mi., 
           Fuel.Cost..USD., 
           Fuel.Volume..gal.) %>%
    group_by(month=paste(as.character(year(Start.Time)), 
                         sprintf('%02d', month(Start.Time)), 
                         sep='-')) %>%
    summarize(miles=sum(Distance..mi.), 
              cost=sum(Fuel.Cost..USD.), 
              gallons=sum(Fuel.Volume..gal.), 
              mpg=miles/gallons, 
              cpg=cost/gallons, 
              cpm=cost/miles) %>% 
    mutate(datasource='Automatic')

mdf <- mileage %>%
    group_by(month=paste(as.character(year(Timestamp)), 
                         sprintf('%02d', month(Timestamp)), 
                         sep='-')) %>%
    summarize(miles=sum(diff.miles), 
              gallons=sum(Gallons), 
              cost=sum(Price), 
              mpg=miles/gallons, 
              cpg=cost/gallons, 
              cpm=cost/miles) %>%
    mutate(datasource='Mileage records')
```

## The data

I have two datasets. One downloaded from the Automatic Dashboard. Each row is a trip. As mentioned above, there are missing trips in this dataset. 

```{r trips, echo=FALSE}
# Dropping a few columns for, you know, privacy
trips %>% 
    select(-contains('Location'), -Path) %>%
    str()
```

The second is a set of data recorded at each fill-up. 

```{r mileage, echo=FALSE}
tail(mileage)
```

## Overall accuracy

Generally, Automatic appears to be accurate when it actually registers a trip. 

```{r overall, echo=FALSE}
# Monthly comparison 
compdf <- rbind(mdf, tdf) %>%
    arrange(month) %>%
    na.omit() %>%
    filter(month>'2014-08')

# Rolled up comparison
rollupdf <- compdf %>%
    group_by(datasource) %>%
    summarize(miles=sum(miles), 
              gallons=sum(gallons), 
              cost=sum(cost), 
              mpg=miles/gallons, 
              cpg=cost/gallons, 
              cpm=cost/miles) 
```

| Measure           | Automatic Data | Mileage Data |
| ------------------|---------------:| ------------:|
| Miles driven      | `r format(rollupdf$miles[1], digits=0, big.mark=',')` | `r format(rollupdf$miles[2], digits=0, big.mark=',')` |
| Gallons consumed  | `r format(rollupdf$gallons[1], digits=0)` | `r format(rollupdf$gallons[2], digits=0)` |
| Fuel cost | `r sprintf('$ %s', format(rollupdf$cost[1], digits=0, big.mark=','))` | `r sprintf('$ %s', format(rollupdf$cost[2], digits=0, big.mark=','))` |
| Miles per gallon | `r format(rollupdf$mpg[1], digits=3)` | `r format(rollupdf$mpg[2], digits=3)` |
| Price per gallon | `r sprintf('$ %s', format(rollupdf$cpg[1], digits=4))` | `r sprintf('$ %s', format(rollupdf$cpg[2], digits=4))` |

Here you can see that, while total mileage and fuel consumption are off, the price per gallon and gas mileage are pretty close to actuals.

## Monthly analysis

Even over time, the Automatic rate data seems accurate. The total monthly mileage tracks well between actuals and Automatic's calculations. I suspect the major month-to-month differences are a result of missed trips and timing mismatches between the trips data and my fill-up mileage data.

```{r comparison, echo=FALSE}
compdf %>%
    ggplot(aes(x=month, y=miles, group=datasource)) + 
    geom_line(aes(color=datasource)) + 
    scale_y_continuous(limits=c(0,2500)) + 
    xlab('Month') + 
    ylab('Total miles') + 
    labs(color='Data source')
```

Rate data (mpg and cost per gallon) show even better accuracy.

```{r mpg, echo=FALSE}
compdf %>%
    ggplot(aes(x=month, y=mpg, group=datasource)) + 
    geom_line(aes(color=datasource)) + 
    scale_y_continuous(limits=c(24.5, 29)) + 
    labs(color='Data source') +     
    ylab('Miles per gallon') + 
    xlab('Month')
```

```{r cpg, echo=FALSE}
compdf %>%
    ggplot(aes(x=month, y=cpg, group=datasource)) + 
    geom_line(aes(color=datasource)) + 
    labs(color='Data source') + 
    scale_y_continuous(limits=c(2.0, 4.0), labels=dollar) + 
    ylab('Price per gallon') + 
    xlab('Month')

```

It's a little disappointing to see the once-plummeting gasoline price level off in February, but it had to end eventually...

## Getting a little more precise

Since trips and fill-ups don't line up perfectly with months, I'd like to compare the trips data directly with each tank of gas. This also gives me a chance to play with time intervals in the [lubridate](http://cran.r-project.org/web/packages/lubridate/index.html) package.

```{r tanks, echo=FALSE}
tank.interval <- data.frame(new_interval(
    mileage$Timestamp[1:nrow(mileage)-1], 
    mileage$Timestamp[2:nrow(mileage)]))

tankdf <- mileage[2:nrow(mileage),] %>%
    select(Gallons, 
           Price, 
           diff.miles) %>% 
    bind_cols(tank.interval)
tankdf <- tankdf %>% 
    mutate(tanknum = rownames(tankdf)) %>%
    rename(tank.interval = 
               new_interval.mileage.Timestamp.1.nrow.mileage....1...mileage.Timestamp.2.nrow.mileage..., 
           gallons = Gallons, 
           cost = Price, 
           miles = diff.miles) %>% 
    mutate(mpg = miles/gallons, 
           cpg = cost/gallons, 
           datasource = 'Mileage records')
#print(tank.interval)
tankdf$tanknum <- as.numeric(tankdf$tanknum)

#trips$trip.interval <- new_interval(trips$Start.Time, trips$End.Time)

getTankId <- function(check.time) {
    tankid <- tankdf$tanknum[check.time %within% tankdf$tank.interval]
    if (length(tankid) == 0) tankid <- NA
    tankid
}

tripdf <- trips %>%
    select(start.time=Start.Time, 
           end.time=End.Time, 
           miles=Distance..mi., 
           duration=Duration..min., 
           cost=Fuel.Cost..USD., 
           gallons=Fuel.Volume..gal.) 
tripdf$tank <- unlist(mapply(getTankId, tripdf$end.time))
tripdf <- tripdf %>% 
    na.omit() %>% 
    group_by(tank) %>% 
    summarize(miles=sum(miles), 
              duration=sum(duration), 
              cost=sum(cost), 
              gallons=sum(gallons)) %>% 
    arrange(tank) %>% 
    mutate(mpg = miles/gallons, 
           cpg = cost/gallons, 
           datasource='Automatic')
```

```{r tankdf, echo=FALSE}
tankdf <- tankdf %>% 
    rename(tank = tanknum) 
tankdf <- merge(tripdf, tankdf, all=T)
tankdf$tank.interval[tankdf$datasource=='Automatic'] <- 
    tankdf$tank.interval[tankdf$datasource=='Mileage records']

tankdf %>% 
    ggplot(aes(x=int_end(tank.interval), y=mpg, group=datasource)) + 
    geom_line(aes(color=datasource)) + 
    labs(color='Data source') + 
    ylab('Miles per gallon') + 
    xlab('')
```

*The big swing you see at the end of November in my mileage records is the result of stopping at a gas station that had some technical problems and I was able to fill my tank only partially. Since accurate estimates of fuel consumption require me to fill the tank each time, this skewed the numbers for this fill-up and the following one.*

```{r cpg2, echo=FALSE}
tankdf %>% 
    ggplot(aes(x=int_end(tank.interval), y=cpg, group=datasource)) + 
    geom_line(aes(color=datasource)) +
    labs(color='Data source') + 
    scale_y_continuous(limits=c(2.1, 4.1), labels=dollar) + 
    ylab('Price per gallon') + 
    xlab('')
```

```{r miles, echo=FALSE}
tankdf %>% 
    ggplot(aes(x=int_end(tank.interval), y=miles, group=datasource)) + 
    geom_line(aes(color=datasource)) +
    labs(color='Data source') + 
    scale_y_continuous(limits=c(0, 400)) +
    ylab('Miles driven') + 
    xlab('')
```

## Conclusion

As you can see in the graphs above, cost per gallon, gas mileage and even total miles driven tracked pretty well with my manually recorded data -- the exceptions coming mostly when AUtomatic missed trips. There are probably other systematic sources of error, too. For instance, I usually pay a slightly higher credit price at the pump, and Automatic might be assuming the lower cash price.
